A general Nullstellensatz for generalised spaces

### 2019-07-10 slides-birmingham2019.tex
Note slides finished

    The asterisks in \emph{geometric$^\star$ sequent} and
    \emph{provability$^\star$} indicate that any infinities used to index
    disjunctions have to be come from the base topos. This restriction is an important
    subtlety, though not vital to this talk. If~$\TT$ is a coherent
    theory, then for coherent sequents there is no difference between
    provability in coherent logic, provability in geometric logic and
    provability$^\star$.
  }

@@

  \jnote{3}{
    While, as stated on slide 1/20, the generic model~$U_\TT$ is a
    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
    believe this fact. That is, the statement~``$U_\TT$ is a
    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
    What is true is the modified statement~``$U_\TT$ is a
    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
  }

>>>>

    The asterisks in \emph{geometric$^\star$ sequent} and
    \emph{provability$^\star$} indicate that any infinities used to index
    disjunctions have to come from the base topos. This restriction is an important
    subtlety, though not vital to this talk. If~$\TT$ is a coherent
    theory, then for coherent sequents there is no difference between
    provability in coherent logic, provability in geometric logic and
    provability$^\star$.
  }

@@

  \jnote{3}{
    While, as stated on slide 1/10, the generic model~$U_\TT$ is a
    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
    believe this fact. That is, the statement~``$U_\TT$ is a
    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
    What is true is the modified statement~``$U_\TT$ is a
    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
  }

### 2019-07-10 slides-edinburgh2019.tex
Note slides finished

%\usepackage{pgfpages}
%\setbeameroption{show notes on second screen}
%\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{}

@@

  Ingo Blechschmidt \\
  Università di Verona
  \par
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight]{arthurs-seat-4}\end{minipage}}
\begin{frame}[plain]%
  \vspace*{\paperheight}\vspace*{-2.35cm}\hspace*{-1cm}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-7}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-0}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-5}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
\end{frame}}
\addtocounter{framenumber}{-1}

@@

\begin{frame}{The mystery of nongeometric sequents}
  \jnote{1}{A \emph{geometric sequent} is a syntactical expression of the
  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
  \emph{geometric formulas}. Often the variable context is abbreviated to~$\vec
  x \? \vec X$ or even just~$\vec x$. Such a sequent is read as ``in the
  context of variables~$\vec x$, $\varphi$ entails~$\psi$''.

  Geometric formulas are built from atomic propositions (using equality or the
  relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
  contain~$\neg$, $\Rightarrow$, $\forall$.

  There is a notion of a \emph{model} of a geometric theory in
  a given topos. For instance, a ring in the usual sense is a model of the
  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.

@@

  For instance, if we want to verify a geometric sequent for all local rings,
  we may freely use the displayed field axiom.
  Hence one reason why these nongeometric sequents are interesting is because
  they provide us with new reduction strategies (``without loss of
  generality'').}

@@

  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
  an equational theory), the \emph{term algebra} (the set of terms in the
  empty context modulo provable equality) is a model of~$\TT$. While such
  models do enjoy some nice categorical properties, they are in general
  \emph{not} the generic model.

@@

  In general, the generic model cannot be realised as a set-based model (with
  a set for each sort, a map for each function symbol and so on). Sets are too
  constant for this purpose; the flexibility of sheaves (``variable sets'') is
  required: The generic model lives in the topos of set-valued
  sheaves over~$\C_\TT$.}

  \jnote{2}{A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
  equivalence class (modulo provable equivalence) of a geometric
  formula~$\theta$ such that~$\TT$ proves
  \begin{enumerate}
    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
    \vec y} \varphi \wedge \psi)$
    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
    \theta)$
    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
  \end{enumerate}

@@

    The internal language can be implemented by the \emph{Kripke--Joyal semantics},
    a translation procedure which converts formulas of the internal language
    into external statements about the objects and morphisms of~$\E$. The slide
    displays some of the translation rules in the case that~$\E$ is a
    Grothendieck topos.

    We can actually do mathematics internally because the Kripke--Joyal
    semantics is sound with respect to intuitionistic logic: If~$\E \models
    \varphi$ and if~$\varphi$ intuitionistically entails a further
    formula~$\psi$, then~$\E \models \psi$.

    An instructive special case is provided by the topos~$\Set$, because~$\Set
    \models \varphi$ iff~$\varphi$ holds in the usual mathematical sense.
  }

  \jnote{2}{
    The Kripke--Joyal semantics can be extended to interpret unbounded
    quantification (``for all sets'' as opposed to ``for all elements of the
    particular set~$X$'') and dependent types. The former are for instance
    required to express universal properties (``for all groups'', ``for all
    rings''), and the latter are all over the place, even if their use might
    not be particularly highlighted.

@@

  \jnote{1}{
    The generic object, the generic model of the theory which has exactly one
    sort and no function symbols, relations symbols or axioms, appears to be
    slightly indecisive: On the one hand, up to a double negation, it is a
    subsingleton; on the other hand, it is infinite. This observation is due to
    Carsten Butz and Peter Johnstone
    (\fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying toposes for
    first-order theories}).

@@

\begin{frame}{A different kind of example}
  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).

@@

  Consequence: short and conceptual proof of Grothendieck's generic freeness
  lemma
\end{frame}

%\section{Affine schemes}
%
%\begin{frame}{Affine schemes}
%  \jnote{1}{
%    A ring is \emph{local} iff every invertible sum contains an invertible
%    summand, that is if~$1 \neq 0$ and if~$x + y$ invertible implies~$x$
%    invertible or~$y$ invertible. Assuming the axiom of choice, this elementary
%    definition is equivalent to the textbook definition of a local ring (a ring
%    with exactly one maximal ideal). A ring homomorphism is \emph{local} iff it
%    reflects invertibility.
%
%    The notion of a \emph{filter} in a ring~$A$ is a direct axiomatisation of
%    what classically would be the complement of a prime ideal. The filter
%    axioms are: $1 \in F$, $(xy \in F) \Leftrightarrow ((x \in F) \wedge (y \in
%    F))$, $\neg(0 \in F)$, $(x+y \in F) \Rightarrow ((x \in F) \vee (y \in F))$.
%
%    A \emph{free local ring} over~$A$ is a homomorphism into a local
%    ring~$A'$ such that any homomorphism into a local ring~$R$ factors uniquely
%    over~$A'$ via a local homomorphism.
%
%    For any particular local ring~$A \xrightarrow{f} R$, the
%    localisation~$A[S^{-1}]$ with~$S = f^{-1}[R^\times]$ is a local ring which
%    fits into the displayed diagram. However, in general there is no single
%    choice of~$S$ which would work for any local ring~$R$. Indeed, classically
%    one can show that a free local ring over~$A$ exists if and only if~$A$
%    contains exactly one prime ideal, in which case~$A$ itself is the free
%    local ring.
%  }
%
%  \jnote{2}{
%    If we want a free local ring to exist for any ring~$A$, we have to broaden
%    our notion of existence and embrace rings which live in toposes other
%    than~$\Set$. There is a notion of a homomorphism between rings living in
%    arbitrary toposes, and using this notion one can verify:
%
%    The free local ring over~$A$ can be built in the classifying topos of
%    filters of~$A$, as the localisation (in that topos) of~$A$ at the generic
%    filter. (More precisely, as the localisation of the mirror image of~$A$ in that
%    topos, that is the constant sheaf~$\underline{A}$.)
%
%    The classifying topos of filters of~$A$ coincides with the topos of sheaves
%    over what's called the \emph{spectrum of~$A$} in algebraic geometry, and
%    under this equivalence the free local ring coincides with the structure
%    sheaf of spectrum. In fact the classifying topos serves as a good
%    constructive substitute for the classical spectrum construction,
%    enjoying the expected universal property even if the axiom of choice is not
%    available, which is why it is simply denoted~``$\Spec(A)$'' on the slide.
%
%    (The classifying topos of prime ideals of~$A$ is also interesting; it
%    coincides with the topos of sheaves over the spectrum of~$A$ equipped with
%    the constructible topology.)
%  }
%
%  \jnote{3}{
%    Assuming the Boolean prime ideal theorem, the geometric sequents
%    validated by~$A^\sim$ are easy to describe: They are precisely those which
%    are validated by all the stalks~$A_\ppp$ of~$A$.
%
%    But~$A^\sim$ enjoys further unique properties which are not shared by the
%    stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
%    reasonable construction. Three of these are displayed on the slide.
%    (A ring is \emph{anonymously Noetherian} iff each of its ideals is
%    \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
%    are constructively acceptable for this Noetherian condition.)
%
%    The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
%    close to~$A$, so that information learned about~$A^\sim$ teaches us
%    about~$A$; on the other hand, it enjoys unique properties rendering it
%    simpler than~$A$.
%
%    This balance allows for a simple and conceptually satisfying proof of
%    \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
%    algebraic geometry. Details can be found in
%    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
%    set of slides}.
%  }
%
%  Let~$A$ be a ring. Is there a \hil{free local ring}~$A \to A'$ over~$A$?
%  \begin{columns}[t]
%    \begin{column}{0.4\textwidth}
%      $\xymatrix{
%        A \ar[rd] \ar[rrr]^f &&& {\substack{\phantom{\text{local}}\\\text{\normalsize$R$}\\\text{local}}} \\
%        & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!35]{\text{local}}[rru]
%      }$
%    \end{column}
%
%    \begin{column}{0.50\textwidth}
%      \small\justifying
%      For a fixed ring~$R$, the localisation $A' \defeq A[S^{-1}]$ with $S \defeq
%      f^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
%      \medskip
%
%      Hence we need the \hil{generic filter}.
%    \end{column}
%  \end{columns}
%
%  \pause
%
%  \justifying
%  The free local ring over~$A$ is~$A^\sim \defeq \ul{A}[F^{-1}]$,
%  where~$F$ is the generic filter, living in~$\Spec(A)$, the classifying topos
%  of filters of~$A$.
%
%  \pause
%
%  If~$A$ is reduced ($x^n = 0 \Rightarrow x = 0$):
%
%  \vspace*{-0.8em}
%  \setbeamercolor{block body}{bg=red!30}
%  \setbeamercolor{structure}{fg=purple}
%  \begin{varblock}{\textwidth}{}
%    $A^\sim$ is a \hil{field}: $\forall x\?A^\sim\_ (\neg(\exists y\?A^\sim\_
%    xy = 1) \Rightarrow x = 0)$.
%
%    $A^\sim$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
%    $\forall x,y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y$.
%
%    \mbox{$A^\sim$ is \hil{anonymously Noetherian}.}\\[-1.2em]
%  \end{varblock}
%\end{frame}

@@

\begin{frame}{A systematic source of nongeometricity?}
  \jnote{1-2}{
    Mimicking the synthetic approach to differential geometry, synthetic
    algebraic geometry is a framework for algebraic geometry in which schemes
    can be modelled by plain old sets, morphisms of schemes by plain old maps
    between those sets, group schemes by plain old groups, and so on. Unlike
    its close cousin, it is far less developed; some first steps are outlined
    in Sections~19 and~20 of
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.

@@

  \jnote{2}{
    Marc Bezem, Ulrik Buchholtz and Thierry Coquand answered in their 2017 paper
    \fixedhref{https://arxiv.org/abs/1712.07743}{Syntactic forcing
    models for coherent logic} Gavin Wraith's question in the negative. (As
    Thierry remarked during the talk, this is even if one takes care to phrase
    the question in a way to exclude the trivial counterexamples given by
    instances of the law of excluded middle in the language of~$\TT$.) If the
    answer had been positive, this would have given a neat, if somewhat hard to
    use in practice, characterisation of the formulas validated by the generic model.
  }

@@

    The asterisks in \emph{geometric$^\star$ sequent}
    indicate that any infinities used to index
    disjunctions have to be come from the base topos. This restriction is an important
    subtlety, though not vital to this talk. If~$\TT$ is a coherent
    theory, then for coherent sequents there is no difference between
    provability in coherent logic, provability in geometric logic and
    provability$^\star$.
  }

  \jnote{2}{
    The algebraic Nullstellensatz states that, in some cases, algebraic truths
    are witnessed by explicit \emph{algebraic certificates} -- syntactical
    objects giving a priori reasons for why a given truth is to be expected.

    In the topos-theoretic Nullstellensatz, algebraic truths are replaced by
    arbitrary truths of the generic model, subject only to the condition that
    they can be expressed as a geometric sequent, and algebraic certificates
    are replaced by \emph{logical certificates}: proofs.
  }

  \jnote{3}{
    While, as stated on slide 1/20, the generic model~$U_\TT$ is a
    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
    believe this fact. That is, the statement~``$U_\TT$ is a
    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
    What is true is the modified statement~``$U_\TT$ is a
    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.
  }

@@

    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ is just an
    ordinary geometric theory, the classifying topos of that theory, and then
    externalising the resulting Grothendieck topos results in the classifying
    topos of~$\TT$-homomorphisms. There are two canonical geometric morphisms
    from this topos to~$\Set[\TT]$, the morphism computing the domain and the
    morphism computing the codomain, and the morphism obtained by the
    externalisation procedure is the former.
  }
\end{frame}

@@

  \only<1>{
    \textbf{In the object classifier.}
  %  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
  %  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
  %  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
  %  \{\heartsuit\}$.
    Let~$x,y \? U_\TT$. Assume that~$\neg(x = y)$.
    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$\sigma \defeq (e_x = e_y \vdash \bot)$.
    But this is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim y)$.
  }

@@

  \jnote{1}{
    This slide gives two examples how to use the Nullstellensatz to deduce
    properties of the generic model. A couple of remarks are in order.

@@

    Often when using the Nullstellensatz, we go from an (assumed) truth
    of~$U_\TT$ via provability to another model~$M$
    of~$\ul{\TT}/U_\TT$. That is, we use provability$^\star$ as a (one-way)
    \emph{bridge}:
    \[ \xymatrix{
      & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
      \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
      \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
    } \]
  }
\end{frame}

@@

  \jnote{1}{
    Theorem~1 states that the source provided by the Nullstellensatz is
    exhaustive. The notion of provability$^{\star\star}$ is a strengthening of
    the notion of provability$^\star$, which in turn is a strengthening of the
    ordinary notion of provability in geometric logic. We are using it here
    because while the notion of provability$^\star$ can be expressed in the
    internal language of a topos, it cannot be expressed in intuitionistic
    logic.

@@

    Theorems~3 and~4 generalise the Nullstellensatz to the
    higher-order setting. The map~$\mathrm{ev}$ maps (the equivalence class
    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.

@@

    Theorem~3 is a relativisation of Olivia Caramello's completeness theorem,
    Theorem~2.4(ii) in her paper
    \fixedhref{https://arxiv.org/abs/0906.3061}{Universal models and
    definability}. The passage from the external to the internal phrasing
    requires going from~$\TT$ to~$\ul{\TT}/U_\TT$.
  }

  \jnote{2}{
    Plans for the future include:
    \begin{itemize}\justifying
      \item Developing an Agda library for dealing with the internal language
      of toposes and related kinds of categories, employing Agda's
      meta-programming facilities; with such a library at hand,
      formalising the Nullstellensatz in Agda.
      \item Exploring Nullstellensatz-style results for arithmetic universes.
      \item Applying the Nullstellensatz in constructive algebra and algebraic
      geometry, along the lines of generic freeness and synthetic algebraic
      geometry. Most of the toposes in geometric use are actually uncharted
      territory from a logical point of view.
    \end{itemize}
  }
\end{frame}}

>>>>

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}

@@

  Ingo Blechschmidt \\
  Università di Verona
  \par

  \jnote{1}{\textbf{This talk in a nutshell:}

  There is a certain ring, the \emph{generic ring}, which is special in that
  it is \emph{conservative}: It has exactly those properties which are shared
  by any ring whatsoever. One could argue that it is this ring which we're
  implicitly referring to when we utter the phrase ``Let~$R$ be a ring.''.

  A surprising observation is that the generic ring is a field. This is not a
  contradiction because conservativity only pertains to properties which can be
  put in a certain syntactic form, dubbed \emph{geometric sequents}, and the
  field condition is not of that form.

  As a consequence, when verifying a geometric sequent for all rings, we may
  freely assume the field condition: Our proof will apply to the generic ring,
  because this ring happens to be a field, and its conclusion will spread to
  any ring whatsoever thanks to its syntactic form.

  An analogous story can be told for the generic group, the generic vector
  space and quite generally for the generic model of any \emph{geometric
  theory}. It is an interesting task to discover properties of such
  generic models. This talk presents a source of such properties, the
  \emph{topos-theoretic Nullstellensatz}, which is
  universal in the sense that any such property is a logical consequence of the
  Nullstellensatz.}
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 5.1cm 0,clip]{arthurs-seat-4}\end{minipage}}
\begin{frame}[plain]%
  \vspace*{\paperheight}\vspace*{-2.35cm}\hspace*{-1cm}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-7}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-0}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-5}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
  \jnote{1}{This talk owes a special debt to Alexander Oldenziel, who
  repeatedly suggested that the topic of this talk should be studied.

  The photos on the slide are of \emph{Arthur's Seat}, a beautiful extinct
  volcano in Edinburgh. It gives an excellent view of the city and is very much
  recommended for a hiking tour.
  }%
\end{frame}}
\addtocounter{framenumber}{-1}

@@

\begin{frame}{The mystery of nongeometric sequents}
  \jnote{1}{A \emph{geometric sequent} is a syntactic expression of the
  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
  \emph{geometric formulas}. Such a sequent is read as ``in the
  context~$x_1\?X_1,\ldots,x_n\?X_n$, $\varphi$ entails $\psi$''.
  Geometric formulas are built from atomic propositions (equality and the
  available relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
  contain~$\neg$, $\Rightarrow$, $\forall$.

  There is a notion of a \emph{model} of a geometric theory in
  any given topos. For instance, a ring in the usual sense is a model of the
  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.

@@

  For instance, if we want to verify a geometric sequent for all local rings,
  we may freely use the displayed field axiom.
  Hence one reason why these nongeometric sequents are interesting is because
  they provide us with new reduction strategies (proofs by ``without loss of
  generality'').}

@@

  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
  an equational theory), the \emph{term algebra} (the set of terms in the
  empty context modulo provable equality) is a model of~$\TT$. While the
  term algebra does enjoy some nice categorical properties, it is in general
  \emph{not} the generic model.

@@

  In general, the generic model cannot be realised as a set-based model (with
  a set for each sort, a map for each function symbol and so on). Sets are too
  constant for this purpose; the flexibility of sheaves (``variable sets'') is
  required: The generic model lives in the topos of set-valued
  sheaves over~$\C_\TT$, called the \emph{classifying topos} of~$\TT$.}

  \jnote{2}{The notation in the definition of the objects of~$\C_\TT$ is meant
  to be suggestive, but should not be taken literally. The objects are
  syntactic expressions consisting of a context and a geometric formula in that
  context.

  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
  equivalence class (modulo provable equivalence) of a geometric
  formula~$\theta$ such that~$\TT$ proves
  \begin{enumerate}
    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
    \vec y} \varphi \wedge \psi)$
    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
    \theta)$
    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
  \end{enumerate}

@@

    The internal language can be implemented by the \emph{Kripke--Joyal semantics},
    a translation procedure which converts formulas of the internal language
    into external statements about the objects and morphisms of~$\E$. The slide
    displays some of the translation rules in the case that~$\E$ is a
    Grothendieck topos.
  }

  \jnote{2}{
    We can actually \emph{do mathematics} internally to a topos because the
    Kripke--Joyal semantics is sound with respect to intuitionistic logic:
    If~$\E \models \varphi$ and if~$\varphi$ intuitionistically entails a
    further formula~$\psi$, then~$\E \models \psi$.

    The Kripke--Joyal semantics can be extended to interpret unbounded
    quantification (``for all sets'' as opposed to ``for all elements of the
    particular set~$X$'') and dependent types. The former are for instance
    required to express universal properties (``for all groups'', ``for all
    rings''), and the latter are all over the place, even if their use might
    not be particularly highlighted.

@@

  \jnote{1}{
    The generic object, the generic model of the theory which has exactly one
    sort and no function symbols, relations symbols or axioms, is
    slightly indecisive: On the one hand, up to a double negation, it is a
    subsingleton; on the other hand, it is infinite. This observation is due to
    Carsten Butz and Peter Johnstone
    (\fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying toposes for
    first-order theories}).

@@

\begin{frame}{An application in commutative algebra}
  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).

@@

  This observation unlocks a short and conceptual proof of Grothendieck's
  generic freeness lemma.

  \jnote{1}{A \emph{local localisation} of a ring~$A$ is a
  localisation~$A[S^{-1}]$ which is a local ring. The classifying topos of
  local localisations of~$A$ is none other than the topos of sheaves over the
  spectrum of~$A$, as studied in algebraic geometry; and the generic model is
  also known as the structure sheaf~$\O_{\operatorname{Spec}(A)}$.

  Assuming the Boolean prime ideal theorem, the geometric sequents
  validated by~$A^\sim$ are easy to describe: They are precisely those which
  are validated by all the stalks~$A_\ppp$ of~$A$.

  But~$A^\sim$ enjoys further unique properties which are not shared by the
  stalks of~$A$, other localisations of~$A$, quotients of~$A$ or indeed any
  reasonable construction. Three of these are displayed on the slide.
  (A ring is \emph{anonymously Noetherian} iff each of its ideals is
  \emph{not not} finitely generated. Textbook proofs of Hilbert's basis theorem
  are constructively valid for this Noetherian condition.)

  The object~$A^\sim$ strikes a fine balance: On the one hand, it is still
  close to~$A$, so that information learned about~$A^\sim$ teaches us
  about~$A$; on the other hand, it enjoys unique properties rendering it
  simpler than~$A$.

  This balance allows for a simple and conceptually satisfying proof of
  \emph{Gro\-then\-dieck's generic freeness lemma}, an important theorem in
  algebraic geometry. Details can be found in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-padova2018.pdf}{this
  set of slides}.}
\end{frame}

@@

\begin{frame}{A systematic source of nongeometricity?}
  \jnote{1-2}{
    Mimicking the synthetic approach to differential geometry, synthetic
    algebraic geometry is a framework for algebraic geometry in which schemes
    are modelled by plain old sets, morphisms of schemes by plain old maps
    between those sets, group schemes by plain old groups, and so on. Unlike
    its close cousin, it is far less developed; some first steps are outlined
    in Sections~19 and~20 of
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these notes}.

@@

  \jnote{2}{
    Marc Bezem, Ulrik Buchholtz and Thierry Coquand answered in their 2017 paper
    \fixedhref{https://arxiv.org/abs/1712.07743}{Syntactic forcing
    models for coherent logic} Gavin Wraith's question in the negative. (As
    Thierry remarked at 6WFTop, this is even if one takes care to phrase
    the question in a way to exclude the trivial counterexamples given by
    instances of the law of excluded middle in the language of~$\TT$.) If the
    answer had been positive, this would have given a neat, if somewhat hard to
    use in practice, characterisation of the formulas validated by the generic model.
  }

@@

    The asterisk in \emph{geometric$^\star$ sequent}
    indicate that any infinities used to index
    disjunctions have to come from the base topos. This restriction is an important
    subtlety, though not vital to this talk. If~$\TT$ is a coherent
    theory, then we can restrict to coherent sequents and thereby vanquish the
    subtlety.
  }

  \jnote{2}{
    The algebraic Nullstellensatz states that, in some cases, algebraic truths
    are witnessed by explicit \emph{algebraic certificates} -- syntactic
    objects giving a priori reasons for why a given truth is to be expected.

    In the topos-theoretic Nullstellensatz, algebraic truths are replaced by
    arbitrary truths of the generic model, subject only to the condition that
    they can be expressed as geometric sequents, and algebraic certificates
    are replaced by \emph{logical certificates}: proofs.
  }

  \jnote{3}{
    While, as stated on slide 1/10, the generic model~$U_\TT$ is a
    conservative~$\TT$-model, the classifying topos~$\Set[\TT]$ does not
    believe this fact. That is, the statement~``$U_\TT$ is a
    conservative~$\ul{\TT}$-model'' is not true internally to~$\Set[\TT]$.
    What is true is the modified statement~``$U_\TT$ is a
    conservative$^\star$~$\ul{\TT}/U_\TT$-model''.

    To be more precise, the following is true: \\
    \mbox{For any geometric sequent~$\sigma$, if~$\Set[\TT] \models
    \speak{$\sigma$ holds for~$U_\TT$}$, then $\TT$ proves~$\sigma$.}

    The following is false in general: \\
    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}$ proves~$\sigma$}$.}

    The following is true: \\
    \mbox{$\Set[\TT] \models \speak{For any geometric$^\star$ sequent~$\sigma$,
    if~$\sigma$ holds for~$U_\TT$, then $\ul{\TT}/U_\TT$ proves~$\sigma$}$.}
  }

@@

    Constructing internally to~$\Set[\TT]$, where~$\ul{\TT}/U_\TT$ looks just like an
    ordinary geometric theory, the classifying topos of that theory, and then
    externalising the resulting Grothendieck topos results in the classifying
    topos of~$\TT$-homomorphisms. There are two canonical geometric morphisms
    from this topos to~$\Set[\TT]$, the morphism computing the domain and the
    morphism computing the codomain, and the morphism obtained by the
    externalisation procedure is the former.
  }
\end{frame}

@@

  \only<1>{
    \textbf{In the object classifier.}
  %  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
  %  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
  %  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
  %  \{\heartsuit\}$.
    Let~$x,y \? U_\TT$. We want to verify~$\neg\neg(x = y)$. Hence assume that~$\neg(x = y)$.
    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$\sigma \defeq (e_x = e_y \vdash \bot)$.
    Thus~$\sigma$ holds in any~$\ul{\TT}/U_\TT$-model.
    But~$\sigma$ is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim
    y)$. Hence~$\bot$.
  }

@@

  \jnote{1}{
    This slide gives an example how to use the Nullstellensatz to deduce
    properties of the generic model. A couple of remarks are in order.

@@

    Often when using the Nullstellensatz, we go from a (perhaps assumed) truth
    of~$U_\TT$ via provability to another model~$M$
    of~$\ul{\TT}/U_\TT$. That is, we use provability as a (one-way)
    \emph{bridge}.
  }
\end{frame}

@@

  \jnote{1}{
    Theorem~1 states that the source provided by the Nullstellensatz is
    exhaustive. It is possible, but not entirely obvious, how to formulate the
    Nullstellensatz as an (infinitary) first-order formula.

@@

    Theorems~3 indicates that the Nullstellensatz can be generalised to the
    higher-order setting.

@@

    Theorem~3 rests on a relativisation of Olivia Caramello's completeness theorem,
    Theorem~2.4(ii) in her paper
    \fixedhref{https://arxiv.org/abs/0906.3061}{Universal models and
    definability}. The passage from the external to the internal phrasing
    requires going from~$\TT$ to~$\ul{\TT}/U_\TT$.
  }

  \jnote{3}{
    Plans for the future include:
    \begin{itemize}\justifying
      \item Developing an Agda library for dealing with the internal language
      of toposes and related kinds of categories, employing Agda's
      meta-programming facilities; with such a library at hand,
      formalising the Nullstellensatz in Agda.
      \item Exploring Nullstellensatz-style results for arithmetic universes.
      \item Applying the Nullstellensatz in constructive algebra and algebraic
      geometry, along the lines of generic freeness and synthetic algebraic
      geometry. Most of the toposes in geometric use are actually uncharted
      territory from a logical point of view. The talk by Matthias Hutzler will
      present new results in that direction.
    \end{itemize}
  }
\end{frame}}
